The Two Dualities of Computation: Negative and Fractional Types

### 2012-03-06 rational.tex
organization
_
related work

Combinators of {{langRev}} can be written using the the operators
described in definition \ref{def:langRev} or via a graphical language
similar in spirit to those developed for GoI
\cite{DBLP:conf/popl/Mackie95} and for category theory
\cite{springerlink:10.1007/978-3-642-12821-94}. Modulo some
conventions and shorthands we will describe here, the wiring diagrams
are a complete and equivalent to the operator based (syntactic)
description of programs.  {{langRev}} combinators expressed in this
graphical language looks like a ``wiring diagrams''. Value take the
form of ``particles'' that flow along the wires. Computation is
expressed by the flow of particles.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{The Field of Rational Types: Intuitions}

%%%%%%%%%%%%%%%%%%
\subsection{Negative Types}
\label{sub:neg}

@@

%%%%%%%%%%%%%%%%%%%%%%
\subsection{Fractional Types}

%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Summary}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{ {{langRevEE}} }

\paragraph*{Observability.} 
Execution of program is defined by {{c : b1 <-> b2}} when evaluated on
input {{v1 : b1}} gives us a value {{v2 : b2}} on
termination. Execution is well defined only if {{b1}} and {{b2}} are
entirely positive types. Consider the program that

@@

@@

@@

@@

(NOTE: Lets use this section here to establish a long of ideas and
intuition. None of these connections have to be formal, but they can
be concrete.)

@@

\paragraph*{Continuations.} The idea of using negative types to model
information flowing backwards, demand for values, continuations, etc. goes
back to at Filinski's thesis. We recall these connections below but we first
note that all these systems are complicated because in all these systems
information can be ignored, destroyed, or duplicated. Clearly the possibility
of erasure of information would mean that our credit card transaction is
incorrect. In our work, information is maintained and hence we have a
guarantee that, in a closed program, the debt must be accounted and paid for.

\paragraph*{Filinski~\cite{Filinski:1989:DCI:648332.755574}.}
In his Masters thesis, Filinski proposes that continuations are a
\emph{declarative} concept. He, furthermore, introduces a symmetric extension
of the $\lambda$-calculus in which values and continuations are treated as
opposites. This is essentially what we are proposing with one fundamental
difference: our underlying language is not the $\lambda$-calculus but a
language of pure isomorphisms in which information is preserved. This shift
of perspective enables us to distill and generalize the duality of values and
continuations: in particular, in the conventional $\lambda$-calculus setting
values and continuations can be erased and duplicated which makes it
difficult to maintain the correspondence between a value and its negative
counterpart. In contrast, in our setting, one can start from the empty type
$0$, introduce a positive value and its negative counterpart, and let each of
these flow in arbitrary ways. The entire framework guarantees that neither
the value nor its negative counterpart will be deleted or duplicated and
hence that, in any closed program, the ``debt'' corresponding to the negative
value is paid off exactly once. The forward and backward executions in our
framework correspond to call-by-value and call-by-name. This duality was
observed by Filinski and others following him but it is particularly clean in
our framework.

\paragraph*{Subtraction} Wadler the reloaded paper, does not consider the
subtraction type because its ``computational interpretation is not
familiar.'' Curien and Herbelin study duality in classical logic, show that
it exchanges call-by-value and call-by-name. They extend classical natural
deduction with subtraction but give it no computational meaning. Crolard (in
the formulae-as-types interpretation of subtractive logic) address the
computational interpretation of subtraction. He explains the type $A-B$ as
the of \emph{coroutines} with a local environment of type $A$ and a
continuation of type $B$. The description is complicated by what is
essentially the desire to enforce linearity constraints so that coroutines
cannot access the local environment of other coroutines. Must cite Selinger
control categories in this context of duality but I am not sure what to say:
it assumes cartesian closed categories for one thing and not symmetric
monoidal ones. Ariola, Herbelin, and Sabry also use subtractive types to
explain delimited continuations.

\paragraph*{Linear Logic} In accounts that are linear, the value and
continuation that comprise the subtractive type need to be constrained to
``stay together.'' This can be achieved by various restrictions. In this work
we have no such constraints, the negative value can flow anywhere. The entire
system guarantees that any closed program would have to account for it. We
don't have to introduce special constraints to achieve that. Zeilberger in
the paper on polarity and the logic of delimited continuations uses polarized
logic: he shows that if positive and negative values are completely symmetric
except that answer types are positive, then the framework accommodates
delimited continuations. But he interprets negative values are control
operators, or as values defined by the shape of their continuations. We
simply interpret values of negative type as values flowing in the ``other''
direction.

\paragraph*{Int Construction.}
For a traced monoidal category {{C}} the Int construction produces a
Compact Closed Category called Int {{C}} \cite{joyal1996traced}.
Further we know that the target of the Int construction is isomorphic
to the target of \G construction of Abramsky \cite{Abramsky96:0} from
Haghverdi. However, note that the {{langRevEE}} is not the same as the
image of the Int construction on {{langRevT}}, since the the later
lacks a multiplicative tensor that distributes over the additive tensor
in Int {{langRevT}}.

\paragraph*{Recursion.} In our previous work we introduce recursive types and
trace operators. This is dangerous here because infinite loops allow us to
prolong paying the debt for as long as we want.

\paragraph*{GoI machines} 
We can now encode the GoI machine of Mackie \cite{Mackie2011,DBLP:conf/popl/Mackie95}.

@@

@@

This highly intuitive and appealing idea was further explored and refined by
many authors~\cite{Griffin:1989:FNC:96709.96714, curien00duality,
  Wadler:2003, DBLP:conf/rta/Wadler05}. Yet, despite its appeal, the duality
between values and continuations

>>>>

Combinators of {{langRev}} can be written using the the operators described
in definition \ref{def:langRev} or via a graphical language similar in spirit
to those developed for GoI~\cite{DBLP:conf/popl/Mackie95} and for category
theory \cite{springerlink:10.1007/978-3-642-12821-94}. Modulo some
conventions and shorthands we will describe here, the wiring diagrams are a
complete and equivalent to the operator based (syntactic) description of
programs.  {{langRev}} combinators expressed in this graphical language looks
like a ``wiring diagrams''. Value take the form of ``particles'' that flow
along the wires. Computation is expressed by the flow of particles.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Negative Types}
\label{sub:neg}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph*{Observability.} 
Execution of program is defined by {{c : b1 <-> b2}} when evaluated on input
{{v1 : b1}} gives us a value {{v2 : b2}} on termination. Execution is well
defined only if {{b1}} and {{b2}} are entirely positive types. This is
similar to the constraint that Zeilberger imposes to explain intuitionistic
polarity and delimited control~\cite{10.1109/LICS.2010.23}.

Consider the program that

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Fractional Types}

@@

For a traced monoidal category {{C}} the Int construction produces a
Compact Closed Category called Int {{C}} \cite{joyal1996traced}.
Further we know that the target of the Int construction is isomorphic
to the target of \G construction of Abramsky \cite{Abramsky96:0} from
Haghverdi. However, note that the {{langRevEE}} is not the same as the
image of the Int construction on {{langRevT}}, since the the later
lacks a multiplicative tensor that distributes over the additive tensor
in Int {{langRevT}}.

@@

We can now encode the GoI machine of
Mackie~\cite{Mackie2011,DBLP:conf/popl/Mackie95}. \amr{The machine uses bang
  which allows arbitrary duplication. Can we really do that???}

@@

\paragraph*{Negation in Classical Linear Logic} 
Filinski~\cite{Filinski92} uses the negative types of linear logic to model
continuations. Reddy~\cite{Reddy91} generalizes this idea by interpreting the
negative types of linear logic as \emph{acceptors}, which are like
continuations in the sense that they take an input and return no
output. Acceptors however are also similar in flavor to logical variables:
they can created and instantiated later once their context of use is
determined. It is clear that, at an intuitive level, acceptors are entities
that combine elements of our negative and fractional types.

\paragraph*{Declarative Continuations.} 
In his Masters thesis~\cite{Filinski:1989:DCI:648332.755574}, Filinski
proposes that continuations are a \emph{declarative} concept. He,
furthermore, introduces a symmetric extension of the $\lambda$-calculus in
which values and continuations are treated as opposites. This is essentially
what we are proposing with one fundamental difference: our underlying
language is not the $\lambda$-calculus but a language of pure isomorphisms in
which information is preserved. This shift of perspective enables us to
distill and generalize the duality of values and continuations: in
particular, in the conventional $\lambda$-calculus setting values and
continuations can be erased and duplicated which makes it difficult to
maintain the correspondence between a value and its negative counterpart. In
contrast, in our setting, one can start from the empty type $0$, introduce a
positive value and its negative counterpart, and let each of these flow in
arbitrary ways. The entire framework guarantees that neither the value nor
its negative counterpart will be deleted or duplicated and hence that, in any
closed program, the ``debt'' corresponding to the negative value is paid off
exactly once. The forward and backward executions in our framework correspond
to call-by-value and call-by-name. This duality was observed by Filinski and
others following him but it is particularly clean in our framework.

\paragraph*{The Duality of Computation.}
Wadler the reloaded paper, does not consider the subtraction type because its
``computational interpretation is not familiar.'' Curien and Herbelin study
duality in classical logic, show that it exchanges call-by-value and
call-by-name. They extend classical natural deduction with subtraction but
give it no computational meaning. Crolard (in the formulae-as-types
interpretation of subtractive logic) address the computational interpretation
of subtraction. He explains the type $A-B$ as the of \emph{coroutines} with a
local environment of type $A$ and a continuation of type $B$. The description
is complicated by what is essentially the desire to enforce linearity
constraints so that coroutines cannot access the local environment of other
coroutines. Must cite Selinger control categories in this context of duality
but I am not sure what to say: it assumes cartesian closed categories for one
thing and not symmetric monoidal ones. Ariola, Herbelin, and Sabry also use
subtractive types to explain delimited continuations.

\paragraph*{Polarized Logic.} Polarized logic is the logic of values and
continuations: each connective can be given a positive or negative polarity,
which can be related to evaluation
order~\cite{10.1109/LICS.2010.23}. Negative types in this work are a mix of
our negative and fractional types: they include a notion of demand which
corresponds to our negative types and a notion of pattern-matching on
continuations (i.e., a notion of constraints on the context) which
corresponds to our fractional types.

\paragraph*{The Lambek-Grishin Calculus.} The ``parsing-as-deduction'' style
of linguistic analysis uses the Lambek-Grishin calculus with the following
types: product, left division, right division, coproduct, right difference,
and left difference~\cite{Bernardi:2010:CSL:1749618.1749689}. The division
and difference types are similar to our types but because the calculus lacks
commutativity and associativity and only has limited notions of
distributivity, each connective needs a left and right version. The
Lambek-Grishin exhibits two notions of symmetry but they are unrelated to our
notions. In particular, the first notion of symmetry expresses commutativity
and the second relates products to sums and divisions to subtractions. In
contract, our two symmetries relate sums to subtractions and products to
divisions.

\newpage

Curien and Herbelin~\cite{Curien:2000} introduce a calculus that exhibits
symmetries between values and continuations and between call-by-value and
call-by-name. The calculus includes the type $A-B$ which is the dual of
implication, i.e., a value of type $A-B$ is a context expecting a function of
type $A \rightarrow B$. Alternatively a value of type $A-B$ is also explained
as a \emph{pair} consisting of a value of type $A$ and a continuation of type
$B$. which whose computational interpretation is left for future work. Curien
and Herbelin informally explain values of types $A-B$ either as a context
expecting a function $A \rightarrow B$ or as a \emph{pair} containing a value
and a continuation. In our interpretation, a value of type $A-B$ is
\emph{either} a value of $A$ or a demand for a value of type $B$.

@@

The idea of using negative types to model information flowing backwards,
demand for values, continuations, etc. goes back to at Filinski's thesis. We
recall these connections below but we first note that all these systems are
complicated because in all these systems information can be ignored,
destroyed, or duplicated. Clearly the possibility of erasure of information
would mean that our credit card transaction is incorrect. In our work,
information is maintained and hence we have a guarantee that, in a closed
program, the debt must be accounted and paid for.

Much of previous work builds on the idea that there is one duality in
computation between values and continuations which manifests itself as a
duality between the call-by-value and call-by-name parameter-passing
mechanisms. The former mechanism focuses on evaluating expressions to values
even if these values are not demanded by the context; the latter focuses on
evaluating expressions to continuations even if these continuations might be
aborted. The idea that a continuation is dual to a value is intuitive but
then one would expect that the sum of two values naturally corresponds to the
sum of two continuations and that the product of two values naturally
corresponds to the product of two continuations.

check and say that our work teases the continuations into a negative part
(which simply demands a value) and a fractional part (which imposes
constraints on how this value will be used). So something like $-1/c$ is
needed to express a conventional continuation. Having two dualities makes the
whole calculus natural and symmetric.

Consider a continuation that takes $x$ and $y$ and swaps them. It can't be
expressed using two conventional continuations because the demand and the way
it is used are entangled together.

In accounts that are linear, the value and continuation that comprise the
subtractive type need to be constrained to ``stay together.'' This can be
achieved by various restrictions. In this work we have no such constraints,
the negative value can flow anywhere. The entire system guarantees that any
closed program would have to account for it. We don't have to introduce
special constraints to achieve that. Zeilberger in the paper on polarity and
the logic of delimited continuations uses polarized logic: he shows that if
positive and negative values are completely symmetric except that answer
types are positive, then the framework accommodates delimited
continuations. But he interprets negative values are control operators, or as
values defined by the shape of their continuations. We simply interpret
values of negative type as values flowing in the ``other'' direction.

@@

In our previous work we introduce recursive types and trace operators. This
is dangerous here because infinite loops allow us to prolong paying the debt
for as long as we want. The algebraic
$\lambda$-calculus~\cite{Vaux:2009:ALC:1630585.1630590} shows that extreme
care is needed when combining negative and fractional types with the full
$\lambda$-calculus (including recursion).

@@

This highly intuitive and appealing idea was further explored and refined by
many authors~\cite{Griffin:1989:FNC:96709.96714, Curien:2000,
  Wadler:2003, DBLP:conf/rta/Wadler05}. Yet, despite its appeal, the duality
between values and continuations
